Causal Inference in one page

### 2015-04-10 causal-bandit/c-bandit.tex
lots of changes - laptop has battery issues

\begin{enumerate}
\item The learner chooses an $I_t \in \set{1,\ldots, K}$ and $J_t \in \set{0,1}$.
\item Then $X_t \in \set{0,1}^K$ is sampled from a product of Bernoulli distributions, $X_{t,i} \sim \Ber(p_i)$ 
\item The learner observes $\tilde X_t \in \set{0,1}^K$, which is defined by 
\eq{
\tilde X_{t,i} = \begin{cases}
X_{t,i} &\text{if } i \neq I_t \\
J_t & \text{otherwise}\,.
\end{cases}
}
\item The learner receives reward $Y_t \sim \Ber(q(\tilde X))$ where $q:\set{0,1}^K \to [0,1]$ is unknown and arbitrary.
\end{enumerate}
The expected reward of taking action $i,j$ is $\mu_{i,j} = \E[q(X)|do(X_i = j)]$. The optimal reward and action are $\mu^*$ and $(i^*,j^*)$ respectively,
where $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and $\mu^* = \mu(i^*,j^*)$. The $n$-step cumulative expected regret is
\eq{
R_n = \E \sum_{t=1}^n \left(\mu^* - \mu_{I_t,J_t}\right).
}

@@

Clearly, this is not quite the right criteria to decide when to switch ...
\subsubsection{Regret on observing for h rounds before switching to UCB}

@@

>>>>

\begin{enumerate}
\item The learner chooses an $I_t \in \set{1,\ldots, K}$ and $J_t \in \set{0,1}$.
\item Then $X_t \in \set{0,1}^K$ is sampled from a product of Bernoulli distributions, $X_{t,i} \sim \Ber(p_i)$ 
\item The learner observes $\tilde X_t \in \set{0,1}^K$, which is defined by 
\eq{
\tilde X_{t,i} = \begin{cases}
X_{t,i} &\text{if } i \neq I_t \\
J_t & \text{otherwise}\,.
\end{cases}
}
\item The learner receives reward $Y_t \sim \Ber(q(\tilde X))$ where $q:\set{0,1}^K \to [0,1]$ is unknown and arbitrary.
\end{enumerate}
The expected reward of taking action $i,j$ is $\mu_{i,j} = \E[q(X)|do(X_i = j)]$. The optimal reward and action are $\mu^*$ and $(i^*,j^*)$ respectively,
where $(i^*,j^*) = \argmax_{i,j} \mu_{i,j}$ and $\mu^* = \mu(i^*,j^*)$. The $n$-step cumulative expected regret is
\eq{
R_n = \E \sum_{t=1}^n \left(\mu^* - \mu_{I_t,J_t}\right).
}
\pagebreak

@@

Clearly, this is not quite the right criteria to decide when to switch ...

What about if I

\subsubsection{Regret on observing for h rounds before switching to UCB}

@@

\subsection{Importance weighted estimators}

Now Hoeffdings bound gives us $ P(\hat{\mu}_i - \mu_{i} > \frac{D}{2}|n_i)$

\eqn{
P(\Delta_{\hat{i^*}} > D) & \leq K \sum_{\boldsymbol{n}}  P(\hat{\mu}_i - \mu_{i} > \frac{D}{2}|n_i)P(\boldsymbol{n}) \\
& = K \sum_{\boldsymbol{n}}  e^{-n_iD^2/2}P(\boldsymbol{n})
}

Where $\boldsymbol{n} = [n_1...n_K]$ is a vector of the number of observations we have for each arm and $P(\boldsymbol{n}) \sim multinomial(\boldsymbol{p},h)$ To approximate this, we break the space of $\boldsymbol{n}$ up into those vectors where $n_i > h/4 \forall i$ and those where $\exists i$ such that $n_i < h/4$.

### 2015-04-10 causal-bandit/regret_bounds.tex
lots of changes - laptop has battery issues

\begin{itemize}
\item $K$ the number of arms
\item $i$ identifies an arm, $i \in \{1...K\}$
\item $i^*$ the arm with the highest true expected reward
\item $I_t$ the arm selected by the algorithm at timestep $t$
\item $\hat{\mu}_i$ an estimator for the expected reward of arm $i$ based on the sample mean
\item $\mu_i$ the true expected reward for an arm $i$
\item $\mu^*$ the true expected reward of  $i^*$ (the best arm)
\item $\Delta_i = \mu^* - \mu_i$ how much worse arm $i$ is than the best arm
\item $T_i(s) = \sum_{t=1}^s \ind{I_t = i}$ the number of times arm $i$ was selected upto timestep $s$
\item $\hat \epsilon_{it} =  (\psi^*)^{-1} \left(\frac{\alpha log(t)}{T_{i}(t-1)}\right)$ an estimate of the uncertainty in the empirical estimator for the expected reward

@@

For the UCB algorithm we make the additional assumption that for each $P(X)$ there exists a convex function $\psi$ such that:

\begin{equation}
\begin{aligned}
log(E[e^{\lambda(X-E[X]}]) & \leq \psi(\lambda)\\
log(E[e^{\lambda(E[X]-X}]) & \leq \psi(\lambda)
\end{aligned}
\end{equation}
This ensures that the moments of the distribution of $X$ are defined and gives us a high probability bound on the how much our sample based estimate of the expected reward can be below the true expected reward. Let  $\hat \mu_{is} = \frac{1}{s}\sum_{t=1}^s X_t$ be the sample average and $\mu_{i} = E[P_i(X)]$ if we select arm $i$ a fixed number of times $s$:

\eqn {
\label{eq:hoeff1}
& P( \mu_{i} - \hat \mu_{is}  > \epsilon) \leq e^{-s\psi^*(\epsilon)} \\
\label{eq:hoeff2}
 \implies & P\left(\mu_{i} - \hat \mu_{is} > (\psi^{*})^{-1}\left( \frac{1}{s}\log\frac{1}{\delta}\right)\right) \leq \delta
}

@@

\eqn {
ucb_{it} & = \hat \mu_{it} + (\psi^{*})^{-1}\left( \frac{\alpha}{T_{i}(t-1)}\log t\right) \text{ , where $\alpha$ is an input parameter} \\
& =  \hat \mu_{it} + \hat \epsilon_{it}
}

@@

Then if $\alpha > 2$,

\begin{equation}
\label{eq:regretBound}
R_n \leq \sum_{i:\Delta i > 0}\left(\frac{\alpha \Delta i \log(n)}{\psi^{*}(\Delta i/2)}+\frac{\alpha}{\alpha-2}\right)
\end{equation}

For $[0,1]$ random variables, with $\psi(\lambda) = \frac{\lambda^2}{8}$ gives $\psi^{*}(\epsilon)=2\epsilon^2$,  $(\psi^{*})^{-1}(x)=\sqrt{\frac{x}{2}}$ and:

\eqn{
ucb_{it} & = \hat \mu_{it} +  \sqrt{\frac{\alpha}{2 T_{i}(t-1)}\log t}
}

\begin{equation}
\label{eq:Bernoulli}
R_n \leq \sum_{i:\Delta i > 0}\left(\frac{2 \alpha \log(n)}{\Delta_i}+\frac{\alpha}{\alpha-2}\right)
\end{equation}

@@

\begin{theorem} If $I_t = i \neq i^*$ at least one of the following statements is true:
\begin{enumerate}
\item The estimated UCB on the best arm, $i^*$, is less than or equal to the actual reward for that arm: $\hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t} \leq \mu^*$
\item The estimated reward for arm $i$ is greater than or equal to the estimated CI higher than the true reward for that arm: $\hat \mu_{it} \geq \mu_i + \hat \epsilon_{it}$
\item The number of times we have selected arm $i$ in previous timesteps, $T_{i}(t-1)$, is less than some bound (that grows logarithmically with $n$). $T_{i}(t-1) < \frac{\alpha log(n)}{\psi^*(\Delta i/2)} \;\;\; \color{red} \longleftarrow \text{feels odd that this grows with $n$, not $t$. Consequence of union bound?}$
\end{enumerate}
\end{theorem}
\begin{proof}
Assume statements 1-3 are all false.
\eq {
3. & \implies T_{i}(t-1) > \frac{\alpha log(n)}{\psi^*(\Delta i/2)} \\
  & \implies \Delta i > 2 (\psi^*)^{-1} \frac{\alpha log(n)}{T_{i}(t-1)} \geq 2 (\psi^*)^{-1} \frac{\alpha log(t)}{T_{i}(t-1)} = 2 \hat \epsilon_{it}\\
  & \implies \Delta i > 2 \hat \epsilon_{it} \\[10pt]
1. & \implies \hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t}  >  \mu^*  = \mu_i+\Delta i > \mu_i +  2 \hat \epsilon_{it}\\
& \implies \hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t}  > \mu_i +  2 \hat \epsilon_{it}\\[10pt]
2. & \implies \hat \mu_{it} < \mu_i + \hat \epsilon_{it} \\
   & \implies \hat \mu_{it} + \hat \epsilon_{it} < \mu_i +2 \hat \epsilon_{it} \\
   & \implies \hat \mu_{it} + \hat \epsilon_{it} <  \hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t} \longleftarrow \text { UCB for arm $i < $ UCB for arm $i^*$, which contradicts $i \neq i^*$}
}

@@

If statements (1) and (2) are both false, then statement (3) places a bound on the number of times we can previously have selected the incorrect arm $i$ in order to select it in this timestep.

Let $\gamma = \ceil{\frac{\alpha \log(n)}{\psi^*(\Delta i/2)}}$ and suppose we had selected arm $i$ in all timesteps until $\gamma$. In the remaining timesteps, we can only select $i$ if statement 3) is false

\eq{
\implies E\left[T_{i}(n)\right] & \leq \gamma + E\left[\sum_{t=1}^{n} \mathds{1} \{I_t = i \text{ and (3) is false} \}\right] \\[3pt]
& \leq \gamma + E\left[\sum_{t=\gamma+1}^{n} \mathds{1} \{\text{(1) or (2) is true} \}\right] \leftarrow \text{since if (3) is false, (1) or (2) must be true}\\[3pt]
& \leq \gamma + \sum_{t=\gamma+1}^{n} \left[ \mathds{P}(\text{(1) is true}) + \mathds{P}(\text{(2) is true})  \}\right] \color{red} \leftarrow \text{Bubeck has $=$ here but are (1) and (2) disjoint?} 
}

\eq {
P(\text{(1) is true})  = & P(\hat{\mu}_{i^*t}+ (\psi^*)^{-1}\left( \frac{\alpha \log{t}}{t} \right) \leq \mu^*)\\
\leq & P(\exists s \in \{1...t\} : \hat{\mu}_{i^*s}+ (\psi^*)^{-1}\left( \frac{\alpha \log{t}}{s} \right) \leq \mu^*) \color{red} \leftarrow \text {to get around the problem that t is random}\\
\leq & \sum_{s=1}^t P\left( \hat{\mu}_{i^*s}+ (\psi^*)^{-1}\left( \frac{\alpha \log{t}}{s} \right) \leq \mu^* \right) \color{red} \leftarrow \text{union bound}\\
\text{ From equation (\ref{eq:hoeff2}) we have:}\\
& P\left( \hat{\mu}_{i^*s}+ (\psi^*)^{-1}\left( \frac{ \log{\frac{1}{\delta}}}{s} \right) \leq \mu^* \right) < \delta\\
\text{Let } \delta = t^{-\alpha} \implies & P\left( \hat{\mu}_{i^*s}+ (\psi^*)^{-1}\left( \frac{\alpha \log{t}}{s} \right)\leq \mu^* \right) < t^{-\alpha}\\
\implies & P(\text{(1) is true}) \leq \sum_{s=1}^t t^{-\alpha} = t*t^{-\alpha} =  t^{1-\alpha}
}

Similarly, $P(\text{(2) is true}) \leq t^{1-\alpha}$ so we have proved:

\eqn {
 E\left[T_{i}(n)\right] & \leq \gamma + \sum_{t=\gamma+1}^n 2t^{1-\alpha}
}

@@

\eqn {
R_n \leq \sum_{i=1}^K \Delta_i  \left( \ceil{\frac{\alpha \log(n)}{\psi^*(\Delta i/2)}} + \sum_{t=\gamma+1}^n 2t^{1-\alpha}\right) 
}

>>>>

\begin{itemize}
\item $K$ the number of arms
\item $i$ identifies an arm, $i \in \{1...K\}$
\item $i^*$ the arm with the highest true expected reward
\item $I_t$ the arm selected by the algorithm at timestep $t$
\item $\hat{\mu}_i$ an estimator for the expected reward of arm $i$ based on the sample mean
\item $\mu_i$ the true expected reward for an arm $i$
\item $\mu^*$ the true expected reward of  $i^*$ (the best arm)
\item $\Delta_i = \mu^* - \mu_i$ how much worse arm $i$ is than the best arm
\item $T_i(s) = \sum_{t=1}^s \ind{I_t = i}$ the number of times arm $i$ was selected upto timestep $s$
\item $\hat \epsilon_{it}$ an estimate of the uncertainty in the empirical estimator for the expected reward

@@

\subsection*{UCB-1}

For simplicity just consider Bernoulli bandits: $P(X_i) \sim Bernoulli(p_i)$. Hoeffding's inequality gives us a high probability bound on the how much our sample based estimate of the expected reward can be below the true expected reward. Let  $\hat \mu_{is} = \frac{1}{s}\sum_{t=1}^s X_t$ be the sample average and $\mu_{i} = E[P_i(X)]$ if we select arm $i$ a fixed number of times $s$:

\eqn {
\label{eq:hoeff1}
& P( \mu_{i} - \hat \mu_{is}  > \epsilon) \leq e^{-2s\epsilon^2} \\
\label{eq:hoeff2}
 \implies & P\left(\mu_{i} - \hat \mu_{is} > \sqrt{\frac{\log(1/\delta)}{2s}}\right) \leq \delta
}

@@

\eqn {
ucb_{it} & = \hat \mu_{it} + \sqrt{\frac{2 \log(t)}{T_i(t)}}\\
& =  \hat \mu_{it} + \hat \epsilon_{it}
}

@@

\eqn{
\label{eq:regretBound}
R_n & \leq \sum_{i:\Delta i > 0}\left(\frac{8\log(n)}{\Delta_i}+(1+\frac{\pi}{3})\Delta_i\right)\\
R_n & < O(\sqrt{K n \log(n)}), \; \text{ provided $n >> K$} 
}

@@

\begin{theorem} If $I_t = i \neq i^*$ at least one of the following statements is true:
\begin{enumerate}
\item The estimated UB on the best arm, $i^*$, is less than or equal to the actual reward for that arm:
\eq{
\hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t} \leq \mu_i+\Delta_i
}
\item The UB on the non-optimal arm is too high: 
\eqn{
\hat \mu_{it}+ \hat \epsilon_{it} \geq \mu_i + 2 \hat \epsilon_{it}
}
\item The uncertainty bound is too wide compared the difference between the payoff of this arm and the optimal one. Since $\hat{\epsilon}_{it}$ is a function of the number of times we have selected arm $i$, this can also be thought of as 'we have not selected arm $i$ enough yet'. 
\eqn{
\Delta_i < 2\hat{\epsilon}_{it}  
}
\end{enumerate}
\end{theorem}
\begin{proof}
Assume statements 1-3 are all false.

\eq{
& UB(i^*) \equiv \hat{\mu}_{i^*t} + \hat{\epsilon}_{i^*t} >  \mu_i+\Delta_i > \mu_i+2\hat{\epsilon}_{it} \\
& UB(i)\equiv \hat{\mu}_{it} + \hat{\epsilon}_{it} < \mu_i+2\hat{\epsilon}_{it}\\
& \implies UB(i) < UB(i^*) \; \text{ which contradicts the statement that we will play arm $i$, $I_t = i$}
}

@@

Suppose we had selected a non-optimal arm $i$ in all timesteps until $\gamma$, where we choose a value $\gamma$ that ensures statement 3 is false. In the remaining time-steps, we can then only select arm $i$ if either statement 1 or statement 2 is true.

\eq{
E\left[T_{i}(n)\right] & \leq \gamma + E\left[\sum_{t=\gamma+1}^{n} \mathds{1} \{\text{(1) or (2) is true} \}\right] \leftarrow \text{since if (3) is false, (1) or (2) must be true}\\[3pt]
& \leq \gamma + \sum_{t=\gamma+1}^{n} \left[ \mathds{P}(\text{(1) is true}) + \mathds{P}(\text{(2) is true})  \}\right]
}

\eq {
P(\text{(1) is true})  = & P(\hat{\mu}_{i^*t} + \sqrt{\frac{2 \log(t)}{T_i(t)}} \leq \mu^*)\\
\leq & P(\exists s \in \{1...t\} : \hat{\mu}_{i^*s}+ \sqrt{\frac{2 \log(t)}{s}} \leq \mu^*) \color{red} \leftarrow \text {to get around the problem that $T_i(t)$ is random}\\
\leq & \sum_{s=1}^t P\left( \hat{\mu}_{i^*s}+ \sqrt{\frac{2 \log(t)}{s}} \leq \mu^* \right) \color{red} \leftarrow \text{union bound}\\
\text{ From equation (\ref{eq:hoeff2}) we have:}\\
& P\left(\mu_{i} - \hat \mu_{is} > \sqrt{\frac{\log(1/\delta)}{2s}}\right) \leq \delta \text{, letting $\delta = t^{-4}$,}\\
\implies & P\left( \hat{\mu}_{i^*s}+\sqrt{\frac{2 \log(t)}{s}}\leq \mu^* \right) < t^{-4}\\
\implies & P(\text{(1) is true}) \leq \sum_{s=1}^t t^{-4} = t*t^{-4} =  t^{-3}
}

Similarly, $P(\text{(2) is true}) \leq t^{-3}$ so we have proved:

\eqn {
 E\left[T_{i}(n)\right] & \leq \gamma + \sum_{t=\gamma+1}^n 2t^{-3}
}

@@

\eqn {
R_n \leq \sum_{i=1}^K \Delta_i  \left(\gamma + \sum_{t=\gamma+1}^n 2t^{-3}\right) 
}

### 2015-04-10 one_page_summary.tex
lots of changes - laptop has battery issues

\begin{document}
\def\ci{\perp\!\!\!\perp}

\subsection*{Causal Inference without the do-opperator}

One way we can encode our assumptions about the state of the world now and how a given intervention modifies that state is with graphical models. Specifically, assume that the data we have is generated from a directed acyclic graph (DAG) and that we know the structure of the graph (although we may not be able to observe all the variables). An arrow $A \rightarrow B$ means $A$ causes $B$.

If we intervene in the system and force a particular variable, $X$ to some value, $x$, the result is a new graph that is identical to the original one except that the conditional probability distribution associated with $X$ is replaced by $X=x$ and any links coming into $X$ are removed (since under the assumption that some external intervention is forcing $X=x$ it will no longer depend on its original causes). The classic example for the question of the causal effect of smoking on lung cancer is shown in figure XXX.

We want to estimate the probability of cancer given smoking after the intervention (ie in the graph $G_X$) using observational data (data generated by the original model $G$). The problem of causal inference in general is to determine if it is possible to estimate the desired conditional probabilities in the graph representing the system after a given intervention from conditional probabilities in the original graph.

This can be determined by repeated applications of Pearl's 3 rules, which we will restate (without the do notation). For simplicity we will only consider the effect of intervention on a single variable, $X$, on a single target $Y$.

\subsubsection*{Rule 1}
 D-separation still applies in the graph $G_X$:

 \begin{equation}
 Y \text{ d-sep } W \text{ given } Z \text{ in } G_X \implies P_{G_X}(Y|Z,W) = P_{G_X}(Y|Z)
 \end{equation}

\subsubsection*{Rule 2}
If, after conditioning on $Z$, $G$ and $G_X$ have exactly the same active trails between $X$ and $Y$ then $P(Y|X,Z)$ will be the same in both graphs $P_G(Y|X,Z) = P_{G_X}(Y|X,Z)$.

Equivalently, we can state this in terms of D-separation by adding an additional variable $\hat{X}$ as a parent of $X$ in the original graph. If $\hat{X}$ d-sep $Y$ given $X$ and $Z$ in $G$ then there are no active trails from $X$ to $Y$ that begin with an arrow into $X$. This means no active trails will be removed in the process of creating graph $G_X$. You can think of the variable $\hat{X}$ as a switch that determines if $X$ takes its value as determined by its other parents or from some external source.

\begin{equation}
Y\text{ d-sep }\hat{X}\text{ given }X,Z \implies P_G(Y|X,Z) = P_{G_X}(Y|X,Z)
\end{equation}

\subsubsection*{Rule 3}
If, after conditioning on $Z$, the only active trails from $X$ to $Y$ in the original graph $G$ begin with an arrow pointing into $X$ then $Y$ will be independent of $X$ given $Z$ after the intervention (since we will have deleted all the arrows going into $X$).

Stated in terms of D-separation:
\begin{equation}
Y \text{ d-sep } \hat{X} \text{ given } Z \implies P_{G_X}(Y|X,Z) = P_{G_X}(Y|Z)
\end{equation}

\subsubsection*{These rules are complete}
We can estimate the conditional distribution of interest in $G_X$ if and only if we can re-write it in terms of observable conditional independences in $G$ via repeated application of these rules.

You might have been hoping for a single, simple graphical criterion for identification - but sadly, to my knowledge, there isn't one. There are some that are sufficient and some that are necessary - but none that are both.

\subsubsection*{The causal effect of smoking}

\subsection*{Causal Inference}
Causal inference covers problems where we want to predict the outcome of an intervention given some known causal structure
\begin{itemize}
\item Assume that the true causal structure is a directed graphical model where $X \rightarrow Y$ means $X$ causes $Y$. We know the structure of the network but may not be able to observe the values of all the variables.
\item Intervening to set $X = x$, denoted $do(X=x)$ in a causal network, has the effect of removing all links coming into the variable $X$ (since its value is no longer determined by its parents) and setting its value to $x$.
\item We can use d-separation to derive the 3 rules of the do-calculus that allow us write queries about interventions in terms of the pre-interventional distribution. 
\item These rules are complete.  If an expression containing do's can be converted into one that does not by repeated application of the rules then you can determine the post-interventional distribution from observational data. If you can't, its not possible without additional assumptions.
\end{itemize}
The rules (simplified to cover only the case where we intervene on a single variable) can all be illustrated by the graph in figure \ref{fig:dorules}. Assume our query is $P(Y|do(X))$. $G^{\dagger}$ is the original graph, $G$, augmented to include a variable $\hat{X}$  that acts as a switch determining if $X$ takes its value as determined by its parents or from an external intervention. $G_{\overline{X}}$ is $G$ after the intervention.

@@

\begin{enumerate}
\item D-separation still applies after an intervention. $Y \ci V|X$ in $G_{\overline{X}}$ $\implies P(Y|do(X),V) = P(Y|do(X))$
\item If the outcome variable is independent of the way in which the intervened on variable takes its value, intervention = observation. $Y \ci \hat{X}|X,Z$ in $G^{\dagger} \implies P(Y|do(X),Z) = P(Y|X,Z)$
\item If there is no direct path from the intervened on variable to the outcome variable, the intervention does not effect the distribution of the outcome variable. $Y \ci \hat{X}|U$ in  $G^{\dagger} \implies P(Y|do(X),U) = P(Y|U)$
\end{enumerate}
In figure \ref{fig:dorules} we can identify the causal effect of $X$ on $Y$ provided we can observe either $Z$ or $U$.

>>>>

\begin{document}
\def\ci{\perp\!\!\!\perp}
\setlength\parindent{0pt}
\subsection*{Causal Inference}
Causal inference covers problems where we want to predict the outcome of an intervention given some known causal structure
\begin{itemize}
\item Assume that the true causal structure is a directed graphical model where $X \rightarrow Y$ means $X$ causes $Y$. We know the structure of the network but may not be able to observe the values of all the variables.
\item Intervening to set $X = x$, denoted $do(X=x)$ in a causal network, has the effect of removing all links coming into the variable $X$ (since its value is no longer determined by its parents) and setting its value to $x$.
\item We can use d-separation to derive the 3 rules of the do-calculus that allow us write queries about interventions in terms of the pre-interventional distribution. 
\item These rules are complete.  If an expression containing do's can be converted into one that does not by repeated application of the rules then you can determine the post-interventional distribution from observational data. If you can't, its not possible without additional assumptions.
\end{itemize}
The rules (simplified to cover only the case where we intervene on a single variable) can all be illustrated by the graph in figure \ref{fig:dorules}. Assume our query is $P(Y|do(X))$. $G^{\dagger}$ is the original graph, $G$, augmented to include a variable $\hat{X}$  that acts as a switch determining if $X$ takes its value as determined by its parents or from an external intervention. $G_{\overline{X}}$ is $G$ after the intervention.

@@

\begin{enumerate}
\item D-separation still applies after an intervention. $Y \ci V|X$ in $G_{\overline{X}}$ $\implies P(Y|do(X),V) = P(Y|do(X))$
\item If the outcome variable is independent of the way in which the intervened on variable takes its value, intervention = observation. $Y \ci \hat{X}|X,Z$ in $G^{\dagger} \implies P(Y|do(X),Z) = P(Y|X,Z)$
\item If there is no direct path from the intervened on variable to the outcome variable, the intervention does not effect the distribution of the outcome variable. $Y \ci \hat{X}|U,Z$ in  $G^{\dagger} \implies P(Y|do(X),U,Z) = P(Y|U,Z)$ (note: we condition on $Z$ because conditioning on $U$ opens the path $\hat{X} \leftarrow Z \rightarrow Y$)
\end{enumerate}

A single application of the rules show we can identify the causal effect of $X$ on $Y$ if we can observe $Z$. By repeated application of the rules the query is also identifiable if we can observe only $U$.
